Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-08-14 manuscript/Reviewer_Reply.md

#  Comments from the SME:

**General comment: I appreciate the authors' efforts to ensure transparency and repeatability of the analysis.**

Thank you.

**l. 20: not all the models are process-based; revise?**

We added changed this to "statistical and processed based models".

**l. 41: could you delete "uncertainty in", or replace with "estimates of"? (Isn't it the timing itself that's important, not the uncertainty in the timing?)**

We deleted "uncertainty in".

**l. 81: fix "quanity".**

Fixed.

**l. 115: why these dates?**

These dates were used in accordance with Crimmins (2017), who used them prior, as budburst is unexpected to be first seen past DOY 172 and DOY 213 for flowers.

**l. 127: phenophase (singular)**

Fixed.

**ll. 210-211. I don't understand this. Why do some seemingly strong relationships (e.g., "naive" panel of Fig. 2; also some panels of Fig. 3) yield negative R-squared values? This probably just reflects my ignorance... but it's likely that if I'm mystified, other readers will be, too! Please explain more clearly how the <0 R-squareds arise.**

**I found the discussion quite insightful, although there are some additional points that should be addressed and a few areas where clarity could be improved (see Reviewer 1's comments)**.

Thank you. Please see the replies to Reviewer 1.

**Also:**
**l. 289: what do you mean by "internal phenology"? internal physiological processes?**

Specifically we meant the date of endodormancy break, but had not used this term prior and did not want to confuse readers unfamiliar with it. We have changed this text to be more descriptive of the Chuine et al. 2016 study results.

**l. 294. "warrant caution" -> "indicate that caution is warranted"**

Fixed.

**l. 305: "the" is repeated.**

Fixed.

**Fig. 4: the labels on the right-hand side ("Metric 1/2") are uninformative. Please use more intuitive labels, e.g. "LTER data/NPN data".**

We replaced the labels to reflect the data evaluated. We also adjusted the methods section slightly to hopefully make this more clear.

# Reviewer(s)' Comments to Author:

## Reviewer: 1

### Comments to the Author
**Comparison of large-scale citizen science data and long-term study data for phenology modeling**

**This is a very practical study and the evaluation of the approach is clear and pretty straightforward to follow. The authors might improve readability a bit by either adding subheads into the Methods & Results or repeatedly using the same terms to refer to the various steps (eg, cross-validation).**

After some discussion, we decided not to add subheadings.

**The most critical thing that I see that’s lacking is that there is no consideration of how decisions made regarding which NPN data to include in the analysis might be impacting model performance. The authors allowed observations with up to 30d between the last reported “no” and the first reported “yes” and chose the midpoint. This could still lead to up to 15d error in the assumed date of “yes” and the true date of “yes”; this variance may in many cases be larger than the signal in the model. Though actually addressing this sensitivity is probably beyond the scope of the analysis, the potential and implications for this error should be discussed explicitly in the paper.**

Gerst et al. 2016 looked at this issue and found that differences in site level means among more restrictive datasets was normally distributed. Thus having a shorter cutoff period here would result equally in less overestimates *and* underestimates. Since the models here are optimzed for RMSE we feel the results would not be affected overall by having a cutoff shorter than 30 days.

Shawn's note:
After talking with Ethan I'll redo most of the analysis with a shorter cutoff date and see how things compare.
also report the distribution of "days between" to see if they're very far off between 30day cutoff and a shorter cutoff.

**The authors lay out two goals for this effort – to determine “inferences about biological processes driving phenology” and to develop predictive phenology models. It may be just me…using the term “inference” in this way confused me.**

**Regarding the predictions that can result from these different models…this is of strong interest to many stakeholder realms. It would be fantastic to see a bit more attention given in the discussion to recommendations for implementation. The authors talk about the tradeoffs between more complex formulations and simpler models and – generally – reasons these differences may come about. It would be great to also see a bit of more practical discussion along the lines of, “if you can tolerate error of 7-14d, then model X may meet your needs, and it’s fine to build the models with either NPN or LTER data. However, if your needs are Y, then you’d be better off doing blah, blah…”… would this be possible?**

note:
This can essentially be done by looking at some of the supplement figures. Will point this out in the discussion.
shoot for 2-3 sentences describing some rules of thumb. with specific of error rates someone can be ok with.

**Thank you for your serious efforts to understand the USA-NPN data and to contribute to the collective understanding of this dataset’s potential as well as shortcomings. I’m so excited to see this sort of work happen! I’m happy to talk with you more regarding this manuscript as well as your other efforts, especially if any of these comments require clarification.**
Sincerely,
Theresa Crimmins, theresa@usanpn.org

### Smaller things

**Please refer to the Network as “USA National Phenology Network” and “USA-NPN” because there are lots of other phenology networks in other countries**

We have added this clarification in several places within the text to avoid any confusion.

**In several places “data” is treated as singular… data are plural**

Fixed.

**L67: the citizen science program run by the USA-NPN that yields the data housed in the National Phenology Database is Nature’s Notebook**
**Thank you for acknowledging the Nature’s Notebook contributors for the data!**

You're welcome. We're happy to credit all data providers, as this study would not have been possible without them.

**Please see the USA-NPN data attribution policy for recommendations for how to properly cite the dataset: https://www.usanpn.org/terms#DataAttribution**

### Abstract
**be sure to indicate that the study is focused on the U.S.**

This is now made clear by specifying that it's the USA NPN.

**L26-29: models performed best when applied to the same data with which they were built – you attribute this to scale…though other things could be at play here, such as local adaptation and species’ sensitivity to different forcing variables varying across the range.**

**L33: you mention that the NPN dataset offers many species… though this is a good point, it is actually irrelevant to the arguments that you are making in this paragraph, that is, if your comparison is focused on only a handful of species.  Oh, now I see you make this point several times in the paper…but I’m still struggling to see how it’s relevant. Is it that you could have built models with many more species in this study based on what NPN offers, but you were limited by what the LTER datasets had to offer?**

Reviewer 2 is correct in that our analysis was limited by species common to the NPN and four LTER datasets. Large scale phenology studies typically employ as many species as are available in a given dataset. Thus we point out the large species pool of the NPN dataset throughout as it is something other researchers should consider when designing their own analysis.

**L34-36: this concluding sentence could be made stronger if you kept the focus on the findings of the present paper – the strengths and limitations of the two datasets and their suitable applications.**

We felt that the strengths and limits of these datasets required more attention in the abstract than could be managed in the final sentence, so we discussed them in the prior sentences primarily and instead looked forward in the final sentence.

### Introduction

**L77: Nature’s Notebook has 1,000s of volunteer participants (www.usanpn.org/data/dashboard)**

Fixed.

**L80: may also want to check out Feldman et al. (2018) – https://doi.org/10.1007/s00484-018-1540-4 for further support on cit sci observer skill**

### Methods

**I think it’s pretty critical to provide your sample sizes for each of the species/phenophases under evaluation for each of the datasets… when I read that you held out 20% of observations for evaluation, I was left wondering, was this 5 points or 50 or 500? You could include this info in Table 1**

We have added details of the sample sizes to the supplement table S1.

**What is the sampling frequency for the LTER sites? This is quite relevant because you chose to use 30d as your cut-off for NPN data - you’ll allow for up to 30d between the last reported “no” and the first reported “yes”. How does this compare to what’s possible in the LTER datasets? It could definitely have an impact on the sensitivity of your results.**

Shawns notes:
put the numbers below in the table or ref. and re-reference them when i talk about reviewer 1's other issue with NPN sampling frequency

sampling freq. from metadata / from data
harvard  3-7 days / 5.6 days
hubbard  7 days / 7.1 days
hjandrews  7 days / 7.6 days
jornada  monthly / 30.5 days

**After reading about the phenology datasets, I expected to read about the temperature data used. I see now that it’s nested within the “Modeling” section… the organization just surprised/confused me.**

The phenological data are the crux of the study so we felt that they deserved special consideration, while the components used to generate the models could be kept within the modeling section. We will gladly change this if the editor feels it is necessary.

**L138-140: Provide some references for your GDD calculation – there are multiple ways to go about calculating GDDs**

The exact calculations are unique to each model and are provided along with references in Table 2. We have clarified this in the text. We also added references for the GDD and Fixed GDD model.

**L143: what is NCD?**

This is the number of chill days (mean daily temperature < 0 degrees C) since Jan. 1 in a given year. We have added the description to the text.

**L145: please provide some references to back up your statement that these formulations are the most common**

**L156: I’m sure you extracted the daily temp at the lat/long of each observation location – maybe just make that a bit clearer**

We have clarified this.

**I don’t think I saw “RMSEA – RMSEB” and “RMSEC-RMSED” referenced anywhere else in the paper**

We removed these specific equations and replaced them with a better description of this analysis.

**L196: what does “scenario A” refer to here?**

This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph slightly and changed the labels on Figure 4 to make this analysis more clear.

### Results

**L231-237: is it possible to quantify these comparisons in any way?**

Suggestion from Ethan:
do a paired t-test here to show distribution is different from 0. Make sure text still matches if a bunch are or are not significantly different than 0.

### Discussion

**L244: are you referring to the temporal depth and/or frequency of sampling when you say “intense”?**

We are referring to both of these as LTER sites typically have a long history of frequent and consistent sampling compared to USA-NPN sites.

**263: “limitations in the sampling of the NPN dataset” – what sort of limitations? Small sample size, frequency of sampling, or something else?**

The frequency of sampling, in a given season and also between seasons, is the primary limit we consider here and we've made this clearer by stating as much.

**L266-268: it would be helpful to add some references for local adaptation and differing sensitivity to phenological drivers across gradients**

We added two references here to highlight the potential variation in phenological requirements across different gradients.

**L270-272: another option could be to develop local or regional models, especially if a national-scale prediction is not needed**

We have briefly included this possibility, as there may indeed be circumstances (population conservation) where the national prediction is unneeded.

**L273: define “model identifiability” and/or provide a reference? I think you are explaining it in the subsequent sentence – if so, just make this more explicit.**

We have made this clearer by switching to a colon instead of a period between the two sentences.

**L276: what do you mean by “these analyses”?**

Maybe change to "our analyses" ?

**L293: not sure what you’re referring to when you say “this previous research”**

We were referencing the two studies cited in this paragraph (Chuine et al. 2016 and Basler 2016). We have changed this to "In combination with the aforementioned studies,..."

**L337: there are many recent studies that have brought together these various datsets; would be worth adding a couple of references**

Shawn's notes on some refs to add:
comparing phenocam + modis: https://doi.org/10.1038/s41598-018-23804-6
comparing modis phenology w/ NPN & Flux towers: 10.1016/j.ecolind.2017.02.024

## Reviewer: 2

### Comments to the Author
Summary:

**While I do think that this manuscript should be ultimately acceptable for publication at Ecology, there are a few major/minor comments and suggestions for which I would like the authors to consider.**

Major Comments and Suggestions:

**First, for NPN species with larger pools/spatial extents of observations, I wonder if would be useful to calibrate the models using subsets somehow stratified by climate and/or latitude. For example, train the models using only warm region (or southern) red maple and test on cold region maples (or vice versa). Perhaps this could further address the issue of model identifiability (as discussed in Lines  259-277). While the oft-mentioned Basler (2016) study certainly had nice spatiotemporal coverage, it did not have nearly the latitudinal/climate variability that NPN affords.**

We discussed this and ultimately decided that it is beyond the scope of the current paper. Although we agree that this approach has the potential to improve the paper, we are unsure of the extent to which this would address the issue of identifiability, which is itself noted as a potential issue rather than a definite one. We also feel that it would require too much space to adequately investigate and discuss, and that the current set of analyses sufficiently accomplish the goals we set out in the introduction. However, we do hope to pursue these confounding factors further in future efforts.

**Second, given the length of time series of the LTER datasets, I think it would be useful to further examine the performance of models based on other criteria such as the amount of interannual variability explained (correlation or R-square) or bias during anomalously warm or cold years. Perhaps the latter might be outside of the overall realm of the paper, but I think it would be informative to know which models capture this type of information which matters in a global change context.**

Shawn's note:
make a R^2 figure for all species/phenophase for the supplement
on the stratified sampling. We thought of this, there's not enough data even for Acer rubrum. Write a sentence or 2 about this in the discussion and how it could be done someday when enought data is collected.

### Minor Comments and Suggestions:

**Line 281: Please verify what is T* (I assume base temperature)**

### 2018-08-14 manuscript/main.tex
double space manuscript
progress as of today on revision
includes notes from meeting with Ethan

% This is the double space command
\renewcommand{\baselinestretch}{1.6}

@@

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data. We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased. Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen. Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

Plant phenology, the timing of recurring biological events such as flowering, plays an important role in ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}. At large scales, uncertainty in the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}. At smaller scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and organisms at higher trophic levels \citep{tylianakis2008}. Plant phenology models that are robust at multiple ecological scales, or deemed appropriate for a particular scale, are needed to better understand and forecast the timing of key biological events.

Many plant phenology studies use intensively collected datasets from a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}. These datasets have regular sampling and large numbers of samples over long periods of time. As a result, the biological and climatic variability at that site is well represented. It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}. For these reasons, data from a single location is not adequate for larger scale phenology modeling. Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, tang2016, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.

Data from citizen science projects is becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}. Because this data is often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams. A relatively new citizen science project started in 2009, The National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}. Data from this project has already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}. Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities. However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}. With hundreds of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high. While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations. This means that the quanity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.

@@

The National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophases combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature. The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 ($NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}. These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.

@@

We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset \citep{prismdata}. We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event. Differential evolution is a global optimization algorithm which uses a population of randomly initialized models to find the set of parameters that minimize the RMSE \citep{storn1997}. Confidence intervals for parameters were obtained by bootstrapping, in which individual models were re-fit 250 times using a random sample, with replacement, of the data. We made predictions by taking the mean DOY estimated from the 250 bootstrapped iterations. A random subset consisting of 20\% of observations from each species and phenophase combination was held out from model fitting for later evaluation.

@@

We also directly evaluated model performance using four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations. Within each of these scenarios, we calculated the RMSE of the held-out observations for each species, phenophase, and model type. Using the RMSE values, we calculated two different metrics to compare the performance of LTER and NPN-derived models on different data types. The first metric focuses on local-scale prediction by comparing the fits of LTER and NPN-derived models on LTER observations: $RMSE_{A} - RMSE_{B}$. The second metric focuses on large-scale prediction by comparing the fits of LTER and NPN-derived models on the NPN data: $RMSE_{C} - RMSE_{D}$. These metrics were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison (scenario A) within each site, not across sites, to focus on local scale prediction when LTER data is available. Absolute RMSE values are provided in the supplement (Fig. S1-S3).

@@

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the Fixed GDD (Fig. 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the NPN dataset, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data is all that is available. Second, spatial variation in phenology requirements could drive these differences, because NPN data integrates over that spatial variation, while LTER data only estimates the phenological requirements for a specific site. In this case, NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}. Third, these differences could result from issues with model identifiability. Since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using these analyses, as will be explained below.

Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Fig. S6-S7). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S4-S5). These results are consistent with research showing that models estimating the starting day of warming accumulation from budbreak time-series failed to accurately infer the internal phenology described in the models \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets. In combination with this previous research, our results warrant caution in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

>>>>

% This is the double space command
\renewcommand{\baselinestretch}{2.0}

@@

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built statistical and process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both NPN and LTER data. We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased. Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen. Predictions for NPN data had the lowest error when using models built from the NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

Plant phenology, the timing of recurring biological events such as flowering, plays an important role in ecological research extending from local to global scales \citep{cleland2007, richardson2013, tang2016}. At large scales the timing of spring leaf out and fall senescence influence the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long-term climate forecasts \citep{richardson2012}. At smaller scales, species-specific responses to temperature and precipitation can alter flower communities \citep{diez2012, caradonna2014, theobald2017} and affect the abundance and richness of both pollinators \citep{ogilvie2017a, ogilvie2017b} and organisms at higher trophic levels \citep{tylianakis2008}. Plant phenology models that are robust at multiple ecological scales, or deemed appropriate for a particular scale, are needed to better understand and forecast the timing of key biological events.

Many plant phenology studies use intensively collected datasets from a single location over a long time-period by a single research group \citep{cook2012, wolkovich2012, iler2013, roberts2015}. These datasets have regular sampling and large numbers of samples over long periods of time. As a result, the biological and climatic variability at that site is well represented. It is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. This lack of transferability can be driven by plasticity in phenology requirements, local adaptation, microclimates, or differences in plant age or population density \citep{kramer1995, diez2012}. For these reasons, data from a single location are not adequate for larger scale phenology modeling. Accurately forecasting phenology at larger scales will require models that account for the full range of variation across a species' range \citep{richardson2013, tang2016, chuine2017}, which will necessitate the use of data sources beyond traditional single-site studies.

Data from citizen science projects are becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}. Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams. A relatively new citizen science project started in 2009, The USA National Phenology Network (NPN), collects phenology observations from volunteers throughout North America and makes the data openly available \citep{npncitation}. Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}. Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the NPN dataset has the potential to meet these needs for North American plant species and communities. However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}. With thousands of participants across North America, the potential for variation among observers in their determination of species identification and dating of phenological events is high. While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants, \citep{fuccillo2015}, observations are sometimes made sporadically across seasons, years, and locations. This means that the quantity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.

@@

The USA National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the NPN and LTER datasets to use in the analysis (Table 1 \& S1).

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2). The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$ \citep{reaumur1735, wang1960, hunter1992}. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 (number of chill days - $NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}. These models are some of the most commonly used in phenology research and serve as a suitable baseline for comparing the NPN and LTER datasets.

@@

We extracted corresponding daily mean temperature for all NPN and LTER observations from the gridded PRISM dataset using the latitude and longitude of the site associated with each observation \citep{prismdata}. We parameterized all models using differential evolution to minimize the root mean square error (RMSE) of the estimated DOY of the phenological event. Differential evolution is a global optimization algorithm which uses a population of randomly initialized models to find the set of parameters that minimize the RMSE \citep{storn1997}. Confidence intervals for parameters were obtained by bootstrapping, in which individual models were re-fit 250 times using a random sample, with replacement, of the data. We made predictions by taking the mean DOY estimated from the 250 bootstrapped iterations. A random subset consisting of 20\% of observations from each species and phenophase combination was held out from model fitting for later evaluation.

@@

We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) NPN-derived models predicting LTER observations, C) LTER-derived models predicting NPN observations, and D) NPN-derived models predicting NPN observations. Using the RMSE values from held out observations, we  compared the  performance of LTER and NPN-derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and NPN-derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the NPN-derived model performed better. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values are provided in the supplement (Fig. S1-S3).

@@

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the Fixed GDD (Fig. 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when NPN data are all that is available. Second, spatial variation in phenology requirements could drive these differences \citep{diez2012, zhang2017}, because NPN data integrate over that spatial variation, while LTER data only estimate the phenological requirements for a specific site. In this case, NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary. Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using these analyses, as will be explained below.

Despite substantial differences in parameter estimates, LTER and NPN-derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both NPN and LTER data (Fig. S6-S7). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S4-S5). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estamates which were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.
